---
title: "Load and Save Hooks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Load and Save Hooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

# Overview

`RunWorkflow()` and `RunWorkflows()` accept an optional `lConfig` object with two data hooks:

- `LoadData`, called before workflow spec validation and before the first step runs.
- `SaveData`, called after the workflow steps finish when `bReturnResult = TRUE`.

Hooks make workflow definitions portable. The workflow can describe what to run, while `lConfig` describes where data should come from and where results should go in a particular environment.

# Example workflow

The examples below use a small workflow that summarizes a measure from an analysis data frame. The workflow does not know how the data is loaded or saved.

```{r}
summarise_measure <- function(adsl, measure) {
  data.frame(
    measure = measure,
    n = nrow(adsl),
    mean = mean(adsl[[measure]])
  )
}

lWorkflow <- list(
  meta = list(
    Type = "demo",
    ID = "height",
    Source = "analysis"
  ),
  steps = list(
    list(
      name = "summarise_measure",
      output = "summary",
      params = list(
        adsl = "adsl",
        measure = "HEIGHTBL"
      )
    )
  )
)

analysis_data <- list(
  analysis = list(
    adsl = data.frame(
      USUBJID = sprintf("SUBJ%03d", 1:4),
      HEIGHTBL = c(160, 172, 168, 181)
    )
  )
)
```

# Inline hooks

For a single project, the simplest configuration is to provide functions directly in `lConfig`.

`LoadData` must accept `lWorkflow`, `lConfig`, and `lData`. It returns the data list that `RunWorkflow()` should use.

`SaveData` must accept `lWorkflow` and `lConfig`. It can read the completed workflow, including `lWorkflow$lData` and `lWorkflow$lResult`.

```{r}
saved_results <- new.env(parent = emptyenv())

lConfig <- list(
  data_store = analysis_data,
  result_store = saved_results,
  LoadData = function(lWorkflow, lConfig, lData) {
    source_name <- lWorkflow$meta$Source
    c(lData, lConfig$data_store[[source_name]])
  },
  SaveData = function(lWorkflow, lConfig) {
    result_name <- paste(lWorkflow$meta$Type, lWorkflow$meta$ID, sep = "_")
    assign(result_name, lWorkflow$lResult, envir = lConfig$result_store)
    invisible(NULL)
  }
)

result <- workr::RunWorkflow(
  lWorkflow = lWorkflow,
  lData = list(),
  lConfig = lConfig
)

result
saved_results$demo_height
```

# Registered providers

When the same loading or saving behavior is shared across workflows or projects, register a named provider once and reference it from `lConfig` by name.

```{r}
workr::register_load_provider(
  "vignette.memory.load",
  function(lWorkflow, lConfig, lData) {
    source_name <- lWorkflow$meta$Source
    c(lData, lConfig$data_store[[source_name]])
  }
)

workr::register_save_provider(
  "vignette.memory.save",
  function(lWorkflow, lConfig) {
    result_name <- paste(lWorkflow$meta$Type, lWorkflow$meta$ID, sep = "_")
    assign(result_name, lWorkflow$lResult, envir = lConfig$result_store)
    invisible(NULL)
  }
)

registered_results <- new.env(parent = emptyenv())

lRegisteredConfig <- list(
  data_store = analysis_data,
  result_store = registered_results,
  LoadData = "vignette.memory.load",
  SaveData = "vignette.memory.save"
)

workr::RunWorkflow(
  lWorkflow = lWorkflow,
  lData = list(),
  lConfig = lRegisteredConfig
)

registered_results$demo_height
```

Provider functions are validated when they are registered. `LoadData` providers must include named formals `lWorkflow`, `lConfig`, and `lData`; `SaveData` providers must include named formals `lWorkflow` and `lConfig`.

# Using hooks with multiple workflows

`RunWorkflows()` passes the same `lConfig` through to each workflow. This lets a project use one loading and saving strategy while each workflow selects the data it needs from `meta`.

```{r}
lWeightWorkflow <- lWorkflow
lWeightWorkflow$meta$ID <- "weight"
lWeightWorkflow$steps[[1]]$params$measure <- "WEIGHTBL"

analysis_data$analysis$adsl$WEIGHTBL <- c(60, 72, 68, 81)

multi_results <- new.env(parent = emptyenv())
lRegisteredConfig$result_store <- multi_results
lRegisteredConfig$data_store <- analysis_data

workr::RunWorkflows(
  lWorkflows = list(lWorkflow, lWeightWorkflow),
  lData = list(),
  lConfig = lRegisteredConfig
)

multi_results$demo_height
multi_results$demo_weight
```

`lConfig` can be either a list or an environment. This is useful when an existing application already stores configuration in an environment, such as a Shiny app or project runtime.
