---
title: "Using the gsm.datasim Load Provider"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the gsm.datasim Load Provider}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Install the optional dependency

The built-in `"gsm.datasim"` provider uses the development version of the optional `{gsm.datasim}` package.

```{r install-gsm-datasim}
pak::pak("Gilead-BioStats/gsm.datasim@dev")
```

# Define a workflow

This workflow summarizes enrolled subjects by simulated site after `RunWorkflow()` loads data with the configured provider.

```{r define-workflow}
library(workr)

example_workflow <- list(
  meta = list(Type = "example", ID = "single_snapshot"),
  steps = list(
    list(
      name = "workr::RunQuery",
      output = "enrolled_subjects_by_site",
      params = list(
        df = "Raw_SUBJ",
        strQuery = paste(
          "SELECT",
          "  invid AS site_id,",
          "  COUNT(*) AS enrolled_subjects,",
          "  AVG(timeonstudy) AS mean_time_on_study",
          "FROM df",
          "WHERE enrollyn = 'Y'",
          "GROUP BY invid",
          "ORDER BY enrolled_subjects DESC",
          sep = "\n"
        )
      )
    )
  )
)
```

# Run a single-snapshot simulation

Set `lConfig$LoadData` to `"gsm.datasim"` and pass provider settings through `lConfig$gsm.datasim`.

```{r run-single-snapshot}
single_snapshot <- workr::RunWorkflow(
  lWorkflow = example_workflow,
  lData = list(caller_value = 11),
  lConfig = list(
    LoadData = "gsm.datasim",
    gsm.datasim = list(
      profile = "standard",
      study_type = "standard",
      participants = 25,
      sites = 4,
      snapshot_count = 1
    )
  ),
  bReturnResult = FALSE
)

sort(names(single_snapshot$lData))
head(single_snapshot$lData$Raw_SUBJ)
head(single_snapshot$lData$enrolled_subjects_by_site)
identical(single_snapshot$lData$Raw_SUBJ, single_snapshot$lData$dfSUBJ)
```

# Run a longitudinal simulation

For longitudinal settings, the provider returns the latest snapshot by default. Use `snapshot` to select a different 1-based snapshot index.

```{r run-longitudinal}
longitudinal_workflow <- example_workflow
longitudinal_workflow$meta$ID <- "longitudinal_case"

latest_snapshot <- workr::RunWorkflow(
  lWorkflow = longitudinal_workflow,
  lConfig = list(
    LoadData = "gsm.datasim",
    gsm.datasim = list(
      study_type = "endpoints",
      months_duration = 2,
      participants = 10,
      sites = 3
    )
  ),
  bReturnResult = FALSE
)

sort(names(latest_snapshot$lData))
head(latest_snapshot$lData$Raw_SUBJ)
```

The same runnable example is available at `inst/examples/gsm-datasim-provider.R`.