---
title: "Incorporating sdtm.oak"
output: html_document
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(rlang)
library(purrr)
library(sdtm.oak)
library(gt)
options("yaml.eval.expr" = TRUE) # needs for admiral -- suggest gsm ecoystem write documentation
```

# Prepare list of data of raw-data
```{r raw data load}
lData <- list(
  dm_raw = read.csv(system.file("raw_data/dm.csv", package = "sdtm.oak")),
  vs_raw =  read.csv(system.file("raw_data/vitals_raw_data.csv", package = "sdtm.oak")),
  study_ct = read.csv(system.file("raw_data/sdtm_ct.csv", package = "sdtm.oak"))
)
```

<details>
<summary>Show Raw DM preview (first 6 rows)</summary>

```{r Show Raw DM, echo = FALSE}
head(lData$dm_raw) %>% gt()
```
</details>

<details>
<summary>Show Raw VS preview (first 6 rows)</summary>
```{r Show Raw VS, echo = FALSE}
head(lData$vs_raw) %>% gt()
```
</details>

<details>
<summary>Show Study CT preview (first 6 rows)</summary>
```{r Show Study Controlled Terminology, echo = FALSE}
head(lData$study_ct) %>% gt()
```
</details>

## Show YAML's of sdtm.oak transformations

::: {.scrollbox}
```{r Show DM SDTM transformation}
#| echo: false

dm_path <- system.file("demo_gsmpharmaverse", "workflows", "1_RAW_TO_SDTM", "DM.yaml", package = "workr")
x <- readLines(dm_path, warn = FALSE)
cat("```yaml\n", paste(x, collapse = "\n"), "\n```\n", sep = "")
```
:::

::: {.scrollbox}
```{r Show VS SDTM transformation}
#| echo: false

vs_path <- system.file("demo_gsmpharmaverse", "workflows", "1_RAW_TO_SDTM", "VS.yaml", package = "workr")
x <- readLines(vs_path, warn = FALSE)
cat("```yaml\n", paste(x, collapse = "\n"), "\n```\n", sep = "")
```
:::

This workflow was designed to mimic and follow this respective `sdtm.oak` [vignette](https://pharmaverse.github.io/sdtm.oak/articles/findings_domain.html). 


# Run Respective Workflows to Arrive at SDTM datasets
```{r SDTM Workflows, message = FALSE}
SDTM_workflows <- workr::MakeWorkflowList(
  strNames = c("VS", "DM"),
  strPath = "demo_gsmpharmaverse/workflows/1_RAW_TO_SDTM/",
  strPackage = "workr"
)
SDTM <- workr::RunWorkflows(lWorkflows = SDTM_workflows, lData = lData)
```

```{r Save SDTM datasets, eval = FALSE}
# Take Results and save them as parquets in a SDTM folder 
# or wherever it needs to be saved, gsm has a load/save configuartion as well
purrr::map2(
  SDTM,
  names(SDTM),
  function(x,y) arrow::write_parquet(x, paste0("demo_gsmpharmaverse/data/SDTM/", y,".parquet")) 
)
```


<details>
<summary>Show SDTM VS (first 6 rows)</summary>
```{r, echo = FALSE}
head(SDTM$SDTM_VS) %>% gt()
```
</details>
